
%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[12pt]{article}
\usepackage{geometry}                
\geometry{letterpaper}                   
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{framed}
\usepackage{listings}
\lstset{language=Octave,
  basicstyle=\footnotesize,
  tabsize=3,
  language=python,
  breaklines=true,
  mathescape=true,
  commentstyle=\color{blue},
  morecomment=[l]{\#}}
\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}
\setromanfont[Mapping=tex-text]{Hoefler Text}
\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Gill Sans}
\setmonofont[Scale=MatchLowercase]{Andale Mono}


\author{Elliott Hauser}
\title{Comp 555 Problem Set 4}
\begin{document}
\maketitle

\section*{Question 1}
1. Consider the following four sequences 
Chimp GTTG
Gorilla  GTCA
Human  ACCA
Orangutan ATTA
Assume the following scoring matrix
A T G C
A 0 2 3 8
T 2 0 1 3
G 3 1 0 3
C 8 3 3 0

a. For these four species, draw any six binary tree topologies. Make 
at least two different tree structures. 
b. Apply Sankoff’s algorithm to the topologies created in (a) to 
determine the weighted parsimony score. What is the minimum parsimony 
score? What is the most plausible topology of the six you created? 
Why?
c. Apply Fitch’s algorithm to the same topologies. Find the sequences 
for the roots for all. Which is the one with the least parsimony 
score. 
d. Do b) and c) give same results. Why or Why not?

\section*{Question 2}
2. Consider the following distance matrix
A B C D E F
A 0 18 15 21 6 16
B 0 23 19 20 24
C 0 26 17 19
D 0 23 27
E 0 18
F 0
a) Is this distance matrix additive ? Why or why not?
b) Design an efficient method to determine “delta” for an iteration. 
c) Find “delta” and i,j,k for the first two iterations.

\section*{Question 3}
3. Book 10.8


4. Book 11.6


5. In problem 4 (book 11.6), for the sequence 112122, what is the 
probability that third 1 came from die D1


\end{document}